Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-09-16 summary.Rmd
forgot to add some citations
fixed figures

```{r}
import::from(magrittr, `%>%`)
library(plot.matrix)
library(ggplot2)

@@

* *Case $k \neq l$*. $P^{(kl)}$ and $A^{(kl)}$ represent edge probabilities and 
edges between communities $k$ and $l$. Note that $P^{(kl)} = (P^{(lk)})^\top$.  
By definition, $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. As in the 
$k = l$ case, we note that the singular value decomposition 
$P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top$ is one-dimensional and 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)}$. (We can also note that the SVD of 
$P^{(lk)} = \sigma_{kl}^2 v^{(kl)} (u^{(kl)})^\top$, i.e., 
$\sigma_{kl} = \sigma_{lk}$, $u^{(kl)} = v^{(lk)}$, and 
$v^{(kl)} = u^{(lk)}$.)  
Now consider the Hermitian dilation 
$M^{(kl)} = 2 \begin{bmatrix} 0 & P^{(kl)} \\ P^{(lk)} & 0 \end{bmatrix}$, 
which is a symmetric $(n_k + n_l) \times (n_k + n_l)$ matrix. It can be shown 
that the eigendecomposition of $M^{(kl)}$ is  
$M^{(kl)} = 
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix} \times 
\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix}^\top$.  
Thus treating $M^{(kl)}$ as the edge probability matrix of a GRDPG, we have 
latent positions in $\mathbb{R}^2$ given by 
$\begin{bmatrix} 
  \sigma_{kl} u^{(kl)} & \sigma_{kl} u^{(kl)} \\ 
  \sigma_{kl} v^{(kl)} & -\sigma_{kl} v^{(kl)} 
\end{bmatrix} = 
\begin{bmatrix} 
  \lambda^{(kl)} & \lambda^{(kl)} \\ 
  \lambda^{(lk)} & -\lambda^{(lk)} 
\end{bmatrix}$.  
Now consider $\hat{M}^{(kl)} = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}$. 
Then $\hat{M}^{(kl)} = M^{(kl)} + E'$ where 
$E' = \begin{bmatrix} 0 & E \\ E^\top & 0 \end{bmatrix}$ and $E$ is the 
$n_k \times n_l$ matrix of independent noise (to generate the Bernoulli entries 
in $A^{(kl)}$.  
Then $\hat{M}^{(kl)}$ is an adjacency matrix drawn from $M^{(kl)}$, so its adjacency 
spectral embedding, given by 
$\begin{bmatrix} 
  \hat{\lambda}^{(kl)} & \hat{\lambda}^{(kl)} \\ 
  \hat{\lambda}^{(lk)} & -\hat{\lambda}^{(lk)} 
\end{bmatrix}$ 
where each $\hat{\lambda}^{(kl)}$ is defined as in Algorithm 4, approximates 
the latent positions of $M^{(kl)}$ by the rate in Theorem 5 of 
\citet{rubindelanchy2017statistical}.  
In this case, the indefinite orthogonal transformation $W_*$ in the GRDPG 
result \cite{rubindelanchy2017statistical} is of the form 
$U^\top \hat{U}$. The eigenvalues of $M$ are distinct since the signature 
for this GRDPG is $(1, 1)$, and $U^\top \hat{U}$ is block diagonal, 
resulting in $W_* \stackrel{P}{\to} I$. Therefore, the adjacency spectral embedding of $\hat{M}^{(kl)}$ is 
a direct estimation of the specific latent positions outlined for $M^{(kl)}$, 
up to sign flip.

@@

```{r}
library(mclust)
import::here(cluster.pabm, .from = '~/dev/pabm-grdpg/functions.R')

@@

ari <- round(fossil::adj.rand.index(z, z.hat) * 100)
```

@@

plot.A <- function(A, labels) {
  # https://www.r-graph-gallery.com/79-levelplot-with-ggplot2.html
  A[order(labels), order(labels)] %>% 
    tibble::as_tibble() %>% 
    tibble::rowid_to_column(var = 'X') %>% 
    tidyr::gather(key = 'Y', value = 'Z', -1) %>% 
    dplyr::mutate(Y = as.numeric(gsub('V', '', Y))) %>% 
    ggplot() + 
    geom_tile(aes(x = X, y = Y, fill = Z)) + 
    coord_fixed() + 
    labs(x = NULL, y = NULL) + 
    theme_void() + 
    theme(legend.position = 'none',
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank()) + 
    scale_x_continuous(expand = c(0, 0)) + 
    scale_y_continuous(expand = c(0, 0)) + 
    scale_fill_gradient(low = 'yellow', high = 'red')
}
```

@@

In the second example, we applied Algorithm 2 to the British MPs Twitter 
network \cite{greene2013producing}, the Political Blogs network and the DBLP n
etwork. For this data analysis, we subsetted the data as described by 
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
networks. Our methods underperformed compared to modularity maximization, 
although performance is comparable for the British MPs and DBLP networks. In 
addition, Algorithm 2's runtime is much lower than that of modularity 
maximization.

@@

```{r mp, cache = TRUE, fig.width = 8, fig.height = 2, fig.cap = 'Adjacency matrices of (from left to right) the British MPs, Political Blogs, and DBLP networks after sorting by the clustering outputted by Algorithm 2.'}
gridExtra::grid.arrange(plot.A(A.bmp, zhat.bmp),
                        plot.A(A.pb, zhat.pb),
                        plot.A(A.dblp, zhat.dblp),
                        ncol = 3)

>>>>

```{r}
import::from(magrittr, `%>%`)
library(plot.matrix)
library(ggplot2)
library(mclust)

@@

* *Case $k \neq l$*. $P^{(kl)}$ and $A^{(kl)}$ represent edge probabilities and 
edges between communities $k$ and $l$. Note that $P^{(kl)} = (P^{(lk)})^\top$.  
By definition, $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. As in the 
$k = l$ case, we note that the singular value decomposition 
$P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top$ is one-dimensional and 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)}$. (We can also note that the SVD of 
$P^{(lk)} = \sigma_{kl}^2 v^{(kl)} (u^{(kl)})^\top$, i.e., 
$\sigma_{kl} = \sigma_{lk}$, $u^{(kl)} = v^{(lk)}$, and 
$v^{(kl)} = u^{(lk)}$.)  
Now consider the Hermitian dilation 
$M^{(kl)} = 2 \begin{bmatrix} 0 & P^{(kl)} \\ P^{(lk)} & 0 \end{bmatrix}$, 
which is a symmetric $(n_k + n_l) \times (n_k + n_l)$ matrix. It can be shown 
that the eigendecomposition of $M^{(kl)}$ is  
$M^{(kl)} = 
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix} \times 
\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix}^\top$.  
Thus treating $M^{(kl)}$ as the edge probability matrix of a GRDPG, we have 
latent positions in $\mathbb{R}^2$ given by 
$\begin{bmatrix} 
  \sigma_{kl} u^{(kl)} & \sigma_{kl} u^{(kl)} \\ 
  \sigma_{kl} v^{(kl)} & -\sigma_{kl} v^{(kl)} 
\end{bmatrix} = 
\begin{bmatrix} 
  \lambda^{(kl)} & \lambda^{(kl)} \\ 
  \lambda^{(lk)} & -\lambda^{(lk)} 
\end{bmatrix}$.  
Now consider $\hat{M}^{(kl)} = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}$. 
Then $\hat{M}^{(kl)} = M^{(kl)} + E'$ where 
$E' = \begin{bmatrix} 0 & E \\ E^\top & 0 \end{bmatrix}$ and $E$ is the 
$n_k \times n_l$ matrix of independent noise (to generate the Bernoulli entries 
in $A^{(kl)}$.  
Then $\hat{M}^{(kl)}$ is an adjacency matrix drawn from $M^{(kl)}$, so its adjacency 
spectral embedding, given by 
$\begin{bmatrix} 
  \hat{\lambda}^{(kl)} & \hat{\lambda}^{(kl)} \\ 
  \hat{\lambda}^{(lk)} & -\hat{\lambda}^{(lk)} 
\end{bmatrix}$ 
where each $\hat{\lambda}^{(kl)}$ is defined as in Algorithm 4, approximates 
the latent positions of $M^{(kl)}$ up to indefinite orthogonal transformation 
by the rate given in Theorem 5 of \citet{rubindelanchy2017statistical}.  
In this case, the indefinite orthogonal transformation $W_*$ in the GRDPG 
result \cite{rubindelanchy2017statistical} is of the form 
$U^\top \hat{U}$. The eigenvalues of $M$ are distinct since the signature 
for this GRDPG is $(1, 1)$, and $U^\top \hat{U}$ is block diagonal, 
resulting in $W_* \stackrel{P}{\to} I$. Therefore, the adjacency spectral embedding of $\hat{M}^{(kl)}$ is 
a direct estimation of the specific latent positions outlined for $M^{(kl)}$, 
up to sign flip.

@@

```{r}
import::here(cluster.pabm, .from = '~/dev/pabm-grdpg/functions.R')

@@

# ari <- round(fossil::adj.rand.index(z, z.hat) * 100)
```

@@

plot.A <- function(A, labels) {
  # https://www.r-graph-gallery.com/79-levelplot-with-ggplot2.html
  A[order(labels), order(labels)] %>% 
    tibble::as_tibble() %>% 
    tibble::rowid_to_column(var = 'X') %>% 
    tidyr::gather(key = 'Y', value = 'Z', -1) %>% 
    dplyr::mutate(Y = rev(as.numeric(gsub('V', '', Y)))) %>% 
    ggplot() + 
    geom_tile(aes(x = X, y = Y, fill = Z)) + 
    coord_fixed() + 
    labs(x = NULL, y = NULL) + 
    theme_void() + 
    theme(legend.position = 'none',
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank()) + 
    scale_x_continuous(expand = c(0, 0)) + 
    scale_y_continuous(expand = c(0, 0)) + 
    scale_fill_gradient(low = 'red', high = 'yellow')
}
```

@@

In the second example, we applied Algorithm 2 to the British MPs Twitter 
network \cite{greene2013producing}, the Political Blogs network 
\cite{10.1145/1134271.1134277}, and the DBLP network \cite{NIPS2009_3855} 
\cite{10.1007/978-3-642-15880-3_42}. For this data analysis, we subsetted the 
data as described by 
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
networks. Our methods underperformed compared to modularity maximization, 
although performance is comparable for the British MPs and DBLP networks. In 
addition, Algorithm 2's runtime is much lower than that of modularity 
maximization.

@@

```{r mp, cache = TRUE, fig.width = 8, fig.height = 2, fig.cap = 'Adjacency matrices of (from left to right) the British MPs, Political Blogs, and DBLP networks after sorting by the clustering outputted by Algorithm 2.', fig.dpi = 25}
gridExtra::grid.arrange(plot.A(A.bmp, zhat.bmp),
                        plot.A(A.pb, zhat.pb),
                        plot.A(A.dblp, zhat.dblp),
                        ncol = 3)
